### 2024-09-21 paper/paper.qmd
Update analysis, data cleaning, and testing scripts, and regenerate paper

---
title: "Analysis of Public Opinion Polls Conducted by the City of Toronto"
subtitle: "Insights on Front Yard Parking, Traffic Calming, and Boulevard Cafes"
author:
  - "Tim Chen"
thanks: "Code and data are available at: https://github.com/timchen0326/sta304_paper_1."
date: today
date-format: long
abstract: "This paper analyzes data from the City of Toronto's public opinion polls on key municipal projects. We use a dataset covering several initiatives, including front yard parking, boulevard cafes, and traffic calming. Our analysis reveals significant variation in voter turnout and approval rates across these different projects. The findings provide insights into community engagement and policy preferences, with potential implications for future urban planning and public consultation processes"
format: 
  pdf:
    toc: true        # Enables Table of Contents
    toc-depth: 2     # Specifies depth level of headings included in the ToC
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
# Install and load the required packages
options(repos = c(CRAN = "https://cran.rstudio.com/"))

data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
```

# Introduction

Public opinion polls are an essential part of local governance, offering insights into how communities perceive proposed municipal projects. In the City of Toronto, polls are regularly conducted on topics such as front yard parking, traffic calming, and the installation of boulevard cafes. These polls not only influence policy decisions but also serve as a proxy for measuring community engagement.

The purpose of this paper is to explore the data collected from these polls and to uncover key trends in public opinion. Specifically, we analyze voter turnout, the proportion of favorable outcomes, and how these vary across different types of projects.

The remainder of this paper is structured as follows. @sec-data provides a detailed description of the dataset used. @sec-results presents the analysis of the data, including graphical representations of the results. Finally, @sec-dis discusses the findings and concludes with potential implications for future polling processes.

# Data {#sec-data}

The dataset used for this analysis was sourced from @opendatatorotno and conducted using @citeR. It contains information on several public polls, including fields such as the number of ballots cast, ballots in favor, ballots opposed, and poll results. The dataset spans a period from 2015 to 2023 and covers various municipal initiatives.

The following figure shows the distribution of the number of ballots cast for different types of municipal applications, providing an overview of public participation.
```{r}
#| label: fig-ballots-cast
#| fig-cap: Distribution of Ballots Cast by Application Type
#| echo: false

data %>%
  ggplot(aes(x = application_for, y = ballots_cast, fill = application_for)) +
  geom_boxplot() +
  labs(x = "Application Type", y = "Number of Ballots Cast", title = "Distribution of Ballots Cast by Application Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

@tbl-poll-results provides a summary of the poll results for different application types, including the number of "In Favour" and "Opposed" results for each type.

```{r}
#| label: tbl-poll-results
#| echo: false
#| tbl-cap: "Summary of Poll Results"
data %>%
  group_by(application_for, poll_result) %>%
  summarise(count = n(), .groups = 'drop') %>%
  knitr::kable(col.names = c("Application Type", "Poll Result", "Count"))

```

The dataset also contains information on whether the response rate was met, which is crucial for validating the results of a poll. The following figure shows the proportion of polls that met the response rate requirement.

```{r}
#| label: fig-response-rate
#| fig-cap: Response Rate Met by Application Type
#| echo: false

data %>%
  ggplot(aes(x = application_for, fill = response_rate_met)) +
  geom_bar(position = "dodge") +
  labs(x = "Application Type", y = "Number of Polls", fill = "Response Rate Met", title = "Response Rate Met by Application Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# Results {#sec-results}

## Voter Turnout
Voter turnout, as measured by the number of ballots cast, varies significantly across different application types. For example, Traffic calming initiatives attracted an average of 200 ballots per poll, while front yard parking attracted an average of 100 ballots, as seen in @fig-ballots-cast, while front yard parking proposals generally receive fewer votes. This suggests that residents are more engaged with policies that directly affect public infrastructure and safety.

## Poll Results
The majority of polls analyzed resulted in a favorable outcome ("In Favour"), particularly for front yard parking and boulevard cafes. As shown in , most applications were supported by the public, although there were notable exceptions in the case of traffic calming initiatives, which received a higher number of "Opposed" votes.

## Response Rate
Meeting the response rate requirement is essential for validating poll results. As shown in @fig-response-rate, response rates were more likely to be met for boulevard cafes and front yard parking polls. This could indicate a higher level of engagement in these areas, possibly due to their direct impact on property use.

# Discussion {#sec-dis}

## Key Findings
One of the most significant findings in this analysis is the variation in voter turnout based on the type of municipal application. Traffic calming initiatives, which have a direct impact on road use and safety, attracted the highest number of ballots cast, indicating strong public interest in infrastructure-related projects.

In contrast, front yard parking polls typically received lower voter turnout, possibly because they are seen as affecting a smaller portion of the population. However, these polls still garnered high levels of support from those who participated.

## Limitations and Future Work
One limitation of this study is the potential for missing data. For instance, certain polls lack complete address information, which could impact the analysis of geographic trends in polling results. Future work could explore more granular patterns in voter engagement by incorporating additional demographic and geographic data.

Additionally, further research could investigate the long-term outcomes of these polls, examining whether the results led to sustained community satisfaction with the implemented policies.

\newpage

\appendix

# Appendix {-}

\newpage

# References

### 2024-09-21 scripts/02-data_cleaning.R

@@ -1,44 +1,59 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Cleans the raw poll data from Polls Conducted by the City dataset
+# Author: [Your Name]
+# Date: [Update to current date]
+# Contact: [Your Email]
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Raw data should be available in 'data/raw_data/unedited_data.csv'
+# Any other information needed? None
 
 #### Workspace setup ####
 library(tidyverse)
+library(janitor)
 
 #### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
+# Read in the raw poll data
+raw_data <- read_csv("data/raw_data/raw_data.csv")
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
+# Clean and select relevant columns
+cleaned_data <- 
+  raw_data %>%
+  # Clean column names
+  clean_names() %>%
+  
+  # Select the columns of interest for analysis
+  select(
+    address, 
+    application_for, 
+    ballots_cast, 
+    ballots_in_favour, 
+    ballots_opposed, 
+    close_date, 
+    poll_result, 
+    response_rate_met
+  ) %>%
+  
+  # Handle missing values by removing rows where address or ballots_cast are missing
+  filter(!is.na(address) & !is.na(ballots_cast)) %>%
+  
+  # Convert data types: dates and ballots
   mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
-  ) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "490",
-                                 "49",
-                                 wing_width_mm)) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "6",
-                                 "60",
[…]

### 2024-09-21 scripts/03-test_data.R

@@ -1,15 +1,39 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Sanity check of the raw poll data
+# Author: [Your Name]
+# Date: [Update to current date]
+# Contact: [Your Email]
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Need to have raw poll data
+# Any other information needed? None.
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+
 
 #### Test data ####
+# Read in the raw data
+data <- read_csv("data/raw_data/raw_data.csv")
+
+# Test for negative numbers in relevant numeric columns
+# Assuming ballots_cast, ballots_in_favour, and ballots_opposed should not have negative values
+test_negative_values <- list(
+  ballots_cast = min(data$BALLOTS_CAST, na.rm = TRUE) >= 0,
+  ballots_in_favour = min(data$BALLOTS_IN_FAVOUR, na.rm = TRUE) >= 0,
+  ballots_opposed = min(data$BALLOTS_OPPOSED, na.rm = TRUE) >= 0
+)
+
+# Display test results for negative values
+test_negative_values
+
+# Test for NAs in critical columns
+# Check if there are any NAs in columns like address, ballots_cast, and poll_result
+test_na_values <- list(
+  address_na = all(!is.na(data$ADDRESS)),
+  ballots_cast_na = all(!is.na(data$BALLOTS_CAST)),
+  poll_result_na = all(!is.na(data$POLL_RESULT))
+)
+
+# Display test results for NAs
+test_na_values

### 2024-09-25 paper/paper.qmd
September 24th, 2024 - Finished All Graphs and planned out all ideas for the new dataset

@@
   pdf:
-    toc: true        # Enables Table of Contents
-    toc-depth: 2     # Specifies depth level of headings included in the ToC
+    toc: true
+    toc-depth: 2    
 number-sections: true
@@
 
-```{r}
-#| include: false
-#| warning: false
-#| message: false
+# Setup and Libraries
 
-library(tidyverse)
-# Install and load the required packages
-options(repos = c(CRAN = "https://cran.rstudio.com/"))
+```{r setup, include=FALSE}
 
-data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
+library(ggplot2)
+library(dplyr)
+library(lubridate)
+library(here)
+library(knitr)
+
+# Load the processed data
+data <- read.csv(here("data/analysis_data/analysis_data.csv"))
+
 ```
@@
 
-Public opinion polls are an essential part of local governance, offering insights into how communities perceive proposed municipal projects.
-In the City of Toronto, polls are regularly conducted on topics such as front yard parking, traffic calming, and the installation of boulevard cafes.
-These polls not only influence policy decisions but also serve as a proxy for measuring community engagement.
-
-The purpose of this paper is to explore the data collected from these polls and to uncover key trends in public opinion.
-Specifically, we analyze voter turnout, the proportion of favorable outcomes, and how these vary across different types of projects.
-
-The remainder of this paper is structured as follows.
-@sec-data provides a detailed description of the dataset used.
-@sec-results presents the analysis of the data, including graphical representations of the results.
-Finally, @sec-dis discusses the findings and concludes with potential implications for future polling processes.
+Homelessness, according to the United Nations, has been one of the most prominent social problems that is taking an increasingly important toll on societal well-being during recent decades.
+The state of being homeless is often understood as individuals' inability to stay in a permanent accommodation.
+As homelessness is a fundamental violation of several human rights, the United Nations points to state's failure to provide adequate and affordable housing as the leading cause of the staggering cases (<https://www.ohchr.org/en/special-procedures/sr-housing/homelessness-and-human-rights>).
+This phenomenon is also observed in Canada, where over 11% of Canadians have reported experiencing some form of homelessness (<https://www.statcan.gc.ca/o1/en/plus/5170-homelessness-how-does-it-happen>).
+Moreover, currently over 235,000 people experience homelessness everyday, according to statistics Canada (<https://www150.statcan.gc.ca/n1/pub/82-003-x/2021001/article/00002-eng.htm>).
+While the Canadian state continues to brush over the severeness of the homeless crisis by restating their initiative and program, which frame the narrative in a extremely positive manner, we would like to investigate into the rapidly increasing homeless population to shed light on the need for a more proactive initiative from the state with housing supply and secure social safety net.
+By analyzing in a time-series manner with further dissection by season and gender type, the analysis of this Canadian social crisis seeks to provide more details and insight into this phenomenon.
 
@@
 
-The dataset used for this analysis was sourced from @opendatatorotno and conducted using @citeR.
-It contains information on several public polls, including fields such as the number of ballots cast, ballots in favor, ballots opposed, and poll results.
-The dataset spans a period from 2015 to 2023 and covers various municipal initiatives.
+## Raw Data
 
-The following figure shows the distribution of the number of ballots cast for different types of municipal applications, providing an overview of public participation.
-```{r}
-#| label: fig-ballots-cast
-#| fig-cap: Distribution of Ballots Cast by Application Type
+The data used in this paper is derived from Open Data Toronto and is accessed through the opendatatoronto library (Gelfand 2022).
+The particular dataset analyzed here focuses on daily shelter and overnight services usage in Toronto, including information about shelter occupancy, capacity, and vacancy (Toronto 2024).
+All data analysis was conducted using R (R Core Team 2023) with the assistance of several packages: tidyverse (Wickham et al.
+2019), here (J.
+B.
+Kirill Müller 2020), dplyr (Hadley Wickham 2023), tibble (R.
+F.
+Kirill Müller, Hadley Wickham 2023), janitor (Sam Firke 2023), ggplot2 (Wickham 2016), and knitr (Xie 2023).
+
+The dataset is published by the City of Toronto and features daily shelter occupancy and capacity information across various shelter types, such as room-based and bed-based programs.
+Data collection occurs at 4 a.m.
+daily, and the dataset includes information on individual shelters' service user counts, the total number of beds or rooms available, and the occupancy rates.
+The data is refreshed regularly, and the particular dataset used in this paper is up-to-date as of April 23, 2024.
+Additionally, since this date, the dataset includes occupancy data from hotel-based shelters operated by the Canadian Red Cross Society for refugee claimants.
+The raw dataset also includes fields such as the shelter group (e.g., adult, family, youth, refugee) and the organization operating the shelter.
+
+## Cleaned Data
+
+```{r, warning=FALSE}
+#| label: tbl-cleaned_data
 #| echo: false
+data_head <- head(data)
 
-data %>%
-  ggplot(aes(x = application_for, y = ballots_cast, fill = application_for)) +
-  geom_boxplot() +
-  labs(x = "Application Type", y = "Number of Ballots Cast", title = "Distribution of Ballots Cast by Application Type") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+data_subset <- data_head[, c("occupancy_date", "service_user_count", "program_model")]
+
+# Create the table
+kable(data_subset, col.names = c('Occupancy Date', 'Service User Count', 'Program Model'), 
+      align = "ccc", caption = "Sample of Cleaned Shelter Occupancy Data")
 
@@
 
-@tbl-poll-results provides a summary of the poll results for different application types, including the number of "In Favour" and "Opposed" results for each type.
+```{r, warning=FALSE}
+#| label: fig-service-user-trend
+#| fig-cap: Distribution of Service User Count Over Time by Date of Occupancy
+#| echo: false
 
-```{r}
-#| label: tbl-poll-results
-#| echo: false
-#| tbl-cap: "Summary of Poll Results"
-data %>%
-  group_by(application_for, poll_result) %>%
-  summarise(count = n(), .groups = 'drop') %>%
-  knitr::kable(col.names = c("Application Type", "Poll Result", "Count"))
+data$occupancy_date <- as.Date(data$occupancy_date, format="%Y-%m-%d")
+
+# Create the scatter plot
+ggplot(data, aes(x = occupancy_date, y = service_user_count)) +
+  geom_jitter(color = "gray", alpha = 0.5, width = 0.2, height = 0.2) +
+  theme_minimal() +
+  labs(x = "Date of Sample Collection", y = "Service User Count") +
+  scale_y_continuous(limits = c(0, 100)) +
+  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
 
@@
 
-The dataset also contains information on whether the response rate was met, which is crucial for validating the results of a poll.
-The following figure shows the proportion of polls that met the response rate requirement.
+## Basic Summary Statistics of the Data
 
-```{r}
-#| label: fig-response-rate
-#| fig-cap: Response Rate Met by Application Type
+To gain a better understanding of the structure of the shelter occupancy data, we first examine the number of observations recorded in each year, as displayed in Table 1.
+Over the period from 2021 to 2024, the data collection appears relatively consistent, with the exception of 2024, which shows a noticeable decrease in the number of observations.
+This reduction in data for 2024 could be a result of incomplete data collection or other external factors that impacted data availability for that year.
+The total number of observations fluctuates slightly between 32,000 to 33,000 for 2021 to 2023, with a significant drop to 22,801 in 2024, hinting that additional observations might still be collected as the year progresses.
+
+Moreover, Table 2 provides key summary statistics on critical variables such as occupancy rate (beds), service user count, and capacity (actual beds).
+On average, the occupancy rate for beds is quite high at 96.18%, with a standard deviation of approximately 9.35%, indicating that most shelters are consistently near full capacity.
+Similarly, the mean service user count stands at 38.41, with a high standard deviation of 35.41, suggesting that there is substantial variability in the number of individuals using the shelters across the dataset.
+Notably, the data for capacity (actual beds) is missing, which may indicate incomplete reporting or issues in data collection regarding shelter bed availability.
+Further analysis is required to understand the implications of this missing data.
+The 'Families' sector was excluded from the final visualization due to a significant portion of missing data in the occupancy rate, which resulted in unreliable service user counts.
+Including this category would have distorted the interpretation of trends across sectors.
+Therefore, it was deemed appropriate to focus on the other sectors that have more complete and consistent data.
+
+```{r, warning=FALSE}
+#| label: tbl-observation-by-year
+#| fig-cap: Service User Count by year
 #| echo: false
 
-data %>%
-  ggplot(aes(x = application_for, fill = response_rate_met)) +
-  geom_bar(position = "dodge") +
-  labs(x = "Application Type", y = "Number of Polls", fill = "Response Rate Met", title = "Response Rate Met by Application Type") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+# Count the number of observations by year
+observations_by_year <- table(data$year)
+# Convert capacity_actual_bed to numeric (if it's not already)
+
+# Create a dataframe for the table
+observations_table <- data.frame(
+  Year = names(observations_by_year),
+  Number_of_Observations = as.vector(observations_by_year)
+)
+
+# Calculate mean and standard deviation for relevant columns
+mean_occupancy_rate <- mean(data$occupancy_rate_beds, na.rm = TRUE)
+sd_occupancy_rate <- sd(data$occupancy_rate_beds, na.rm = TRUE)
+
+mean_service_user_count <- mean(data$service_user_count, na.rm = TRUE)
+sd_service_user_count <- sd(data$service_user_count, na.rm = TRUE)
+
+mean_capacity_actual_bed <- mean(data$capacity_actual_bed, na.rm = TRUE)
+sd_capacity_actual_bed <- sd(data$capacity_actual_bed, na.rm = TRUE)
+
+# Table for observations by year
+kable(observations_table, caption = "Table: Number of observations by year")
+```
+
+```{r, warning=FALSE}
+#| label: tbl-summary-statistics
+#| fig-cap: Summary Statistics
+#| echo: false
+#| 
+summary_stats <- data.frame(
+  Variable = c("Occupancy Rate (Beds)", "Service User Count", "Capacity (Actual Beds)"),
+  Mean = c(mean_occupancy_rate, mean_service_user_count, mean_capacity_actual_bed),
+  SD = c(sd_occupancy_rate, sd_service_user_count, sd_capacity_actual_bed)
+)
+
+kable(summary_stats, caption = "Table: Mean and standard deviation of key variables")
 
@@
 
-## Voter Turnout
-Voter turnout, as measured by the number of ballots cast, varies significantly across different application types.
-For example, Traffic calming initiatives attracted an average of 200 ballots per poll, while front yard parking attracted an average of 100 ballots, as seen in @fig-ballots-cast, while front yard parking proposals generally receive fewer votes.
-This suggests that residents are more engaged with policies that directly affect public infrastructure and safety.
+```{r, warning=FALSE}
+#| label: fig-occupancy-rate
+#| fig-cap: Average Occupancy Rate of Beds by year
+#| echo: false
 
-## Poll Results
-The majority of polls analyzed resulted in a favorable outcome ("In Favour"), particularly for front yard parking and boulevard cafes.
-As shown in , most applications were supported by the public, although there were notable exceptions in the case of traffic calming initiatives, which received a higher number of "Opposed" votes.
+yearly_data <- data %>%
+  group_by(year) %>%
+  summarise(Avg_Occupancy_Rate_Beds = mean(as.numeric(occupancy_rate_beds), na.rm = TRUE), 
+            .groups = "keep")
 
-## Response Rate
-Meeting the response rate requirement is essential for validating poll results.
-As shown in @fig-response-rate, response rates were more likely to be met for boulevard cafes and front yard parking polls.
-This could indicate a higher level of engagement in these areas, possibly due to their direct impact on property use.
+# Plot: Average Occupancy Rate of Beds by year
+ggplot(yearly_data, aes(x = factor(year), y = Avg_Occupancy_Rate_Beds)) +
+  geom_bar(stat = "identity", fill = "skyblue") +
+  geom_text(aes(label = round(Avg_Occupancy_Rate_Beds, 1)), vjust = -0.5, color = "black") +
+  labs(title = "Average Occupancy Rate of Beds by year",
+       x = "year", y = "Average Occupancy Rate (Beds)") +
+  theme_minimal() +
+  coord_cartesian(ylim = c(80, NA))
 
-# Discussion {#sec-dis}
+```
 
-## Key Findings
-One of the most significant findings in this analysis is the variation in voter turnout based on the type of municipal application.
-Traffic calming initiatives, which have a direct impact on road use and safety, attracted the highest number of ballots cast, indicating strong public interest in infrastructure-related projects.
+With the graph that detailed the average occupancy rate of beds from 2021 to 2024, it is evident that there is a increasing need for shelter–which most probably translates to an increasing homeless population in Toronto.
+Although the government initiative claimed that there are more resources allocated to assist......
 
-In contrast, front yard parking polls typically received lower voter turnout, possibly because they are seen as affecting a smaller portion of the population.
-However, these polls still garnered high levels of support from those who participated.
+This result can also hint towards a the maximization of capacity that suggest the continuous need for resources, not a staggering increase, We will need to focus on other trends in data to get to the conclusion on whether there is a lack of resources in solving homelessness in Toronto.
 
-## Limitations and Future Work
-One limitation of this study is the potential for missing data.
-For instance, certain polls lack complete address information, which could impact the analysis of geographic trends in polling results.
-Future work could explore more granular patterns in voter engagement by incorporating additional demographic and geographic data.
+```{r, warning=FALSE}
+#| label: fig-bed-capacity
+#| fig-cap: Actual Bed Capacity by Program Model and year
+#| echo: false
+filtered_data <- data %>%
+  filter(program_model %in% c("Emergency", "Transitional"))
 
-Additionally, further research could investigate the long-term outcomes of these polls, examining whether the results led to sustained community satisfaction with the implemented policies.
+# Group by year, Program Model, and summarize actual bed capacity
+yearly_bed_capacity <- filtered_data %>%
+  group_by(year, program_model) %>%
+  summarise(Total_Bed_Capacity = sum(capacity_actual_bed, na.rm = TRUE), 
+            .groups = "keep")
+
+# Plot: Actual Bed Capacity by Program Model and year
+ggplot(yearly_bed_capacity, aes(x = factor(year), y = Total_Bed_Capacity, fill = program_model)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(title = "Actual Bed Capacity by Program Model and year",
+       x = "year", y = "Actual Bed Capacity") +
+  theme_minimal()
+
+```
+
+The above graph shed light on the actual program model that exist in these shelters, giving us a deeper look into what resources the large homeless population in Toronto actually needs.
+The data is focused on two program models....
+each meaning....
+
+with large disparity between the two types, we see a lack of desire in staying in shelter for an extended period of time, which may express the desire to get out of the homeless state or some undesirable nature of these shelters.
+
+The gap between two capacities also can point to possible policy direction that the government and hone in on, as well as potential academic research on the underlying reason behind the great differences.
+
+Moreover, we also see that the capacity for emergency category grew in size progressively within the years that the data has presented.
+This detail echoed the growing rate of occupancy presented above, as well as affirming the assumption on the growing population of homeless individuals.
+With a growing occupancy rate adding on to the actual growth of capacity, there is a strong possibility that the homeless population is seeing a exponential growth that the state has to focus on solving–with a more assertive policy outlook.
+
+```{r , warning=FALSE}
+#| label: fig-service-user-count
+#| fig-cap: Service User Count Over Months by year
+#| echo: false
+monthly_data <- data %>%
+  group_by(year, month) %>%
+  summarise(service_user_count = sum(service_user_count, na.rm = TRUE), 
+            .groups = "keep")
+
+# Plot: Service User Count Over Months by year
+ggplot(monthly_data, aes(x = month, y = service_user_count, color = factor(year), group = year)) +
+  geom_line(size = 1) +
+  labs(title = "Service User Count Over Months by year",
+       x = "month", y = "Service User Count", color = "Year") +
+  theme_minimal() +
+  scale_x_discrete(limits = month.abb)
+```
+
+After looking into capacities of homeless shelters in Toronto, we will now look into service user count from 2021 to 2024, with a monthly distribution that give us a clearer sense of how the capacities
+
+```{r, warning=FALSE}
+#| label: fig-service-sector
+#| fig-cap: Service User Count by sector for Each year
+#| echo: false
+sector_data <- data %>%
+  group_by(year, sector) %>%
+  summarise(Total_Service_User_Count = sum(service_user_count, na.rm = TRUE), .groups = "drop")
+
+sector_data_filtered <- sector_data %>%
+  filter(sector != "Families")
+
+# Plot: Service User Count by sector for Each year
+ggplot(sector_data_filtered, aes(x = sector, y = Total_Service_User_Count, fill = sector)) +
+  geom_bar(stat = "identity") +
+  labs(title = "Service User Count by sector for Each year", x = "sector", y = "Total Service User Count") +
+  facet_wrap(~ year, ncol = 2) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  coord_cartesian(ylim = c(0, NA))
+
+```
 
@@
 
-# Appendix {-}
+# Appendix {.unnumbered}
 

### 2024-09-25 scripts/00-simulate_data.R

@@ -1,49 +1,37 @@
 #### Preamble ####
-# Purpose: Simulates Polls Conducted by the City dataset
+# Purpose: Simulates data
 # Author: Tim Chen
-# Date: 21 September 2024
+# Date: 19 September 2024
 # Contact: timwt.chen@mail.utoronto.ca
 # Pre-requisites: None
 # Any other information needed? None
 
 #### Workspace setup ####
 library(tidyverse)
+library(lubridate)
 
 #### Simulate data ####
 set.seed(304)
 
-# Define the start and end date for poll dates
-start_date <- as.Date("2015-01-01")
+# Define the start and end dates for random date generation
+start_date <- as.Date("2018-01-01")
 end_date <- as.Date("2023-12-31")
 
-# Number of polls to simulate
-number_of_polls <- 100
+# Number of random dates and observations
+n_dates <- 100
 
-# Simulate address data (fake addresses)
-addresses <- paste0(sample(1:1000, number_of_polls, replace = TRUE), " Example St")
-
-# Define possible application types
-applications <- c("Front Yard Parking", "Sidewalk Construction", "Road Repairs")
-
-# Simulate data with selected columns of interest
-poll_data <- tibble(
-  id = 1:number_of_polls,
-  ADDRESS = sample(addresses, number_of_polls, replace = TRUE),
-  APPLICATION_FOR = sample(applications, number_of_polls, replace = TRUE),
-  BALLOTS_CAST = sample(10:50, number_of_polls, replace = TRUE),
-  BALLOTS_IN_FAVOUR = sample(5:25, number_of_polls, replace = TRUE),
-  BALLOTS_OPPOSED = sample(0:10, number_of_polls, replace = TRUE),
-  CLOSE_DATE = as.Date(
-    runif(
-      n = number_of_polls,
-      min = as.numeric(start_date),
-      max = as.numeric(end_date)
-    ),
-    origin = "1970-01-01"
-  ),
-  POLL_RESULT = sample(c("In Favour", "Opposed"), number_of_polls, replace = TRUE),
-  RESPONSE_RATE_MET = sample(c("Yes", "No"), number_of_polls, replace = TRUE)
+# Create simulated data
+simulated_data <- tibble(
+  OCCUPANCY_DATE = as.Date(runif(n = n_dates, 
+                                 min = as.numeric(start_date), 
+                                 max = as.numeric(end_date)), 
[…]

### 2024-09-25 scripts/01-download_data.R

@@ -14,11 +14,11 @@ library(tidyverse)
 
 #### Download data ####
 # get package
-package <- show_package("7bce9bf4-be5c-4261-af01-abfbc3510309")
+package <- show_package("21c83b32-d5a8-4106-a54f-010dbe49f6f2")
 package
  
 # get all resources for this package
-resources <- list_package_resources("7bce9bf4-be5c-4261-af01-abfbc3510309")
+resources <- list_package_resources("21c83b32-d5a8-4106-a54f-010dbe49f6f2")
  
 # identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
 datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
@@ -32,6 +32,6 @@ data
 #### Save data ####
 # [...UPDATE THIS...]
 # change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(data, "data/raw_data/unedited_data.csv") 
+write_csv(data, "data/raw_data/daily-shelter-overnight-service-occupancy-capacity-2024.csvv") 
 
          

### 2024-09-25 scripts/02-data_cleaning.R

@@ -2,58 +2,67 @@
 # Purpose: Cleans the raw poll data from Polls Conducted by the City dataset
 # Author: [Your Name]
 # Date: [Update to current date]
+#### Workspace setup ####
 # Contact: [Your Email]
 # License: MIT
 # Pre-requisites: Raw data should be available in 'data/raw_data/unedited_data.csv'
 # Any other information needed? None
 
-#### Workspace setup ####
-library(tidyverse)
+#### Workspace setup ###
+# Load necessary libraries
+library(readr)
+
+library(lubridate)
+library(tidyr)
+library(dplyr)
 library(janitor)
 
-#### Clean data ####
-# Read in the raw poll data
-raw_data <- read_csv("data/raw_data/raw_data.csv")
+# Read the CSV file
+df2021 <- read.csv("data/raw_data/daily-shelter-overnight-service-occupancy-capacity-2021.csv")
+df2022 <- read.csv("data/raw_data/daily-shelter-overnight-service-occupancy-capacity-2022.csv")
+df2023 <- read.csv("data/raw_data/daily-shelter-overnight-service-occupancy-capacity-2023.csv")
+df2024 <- read.csv("data/raw_data/daily-shelter-overnight-service-occupancy-capacity-2024.csv")
 
-# Clean and select relevant columns
-cleaned_data <- 
-  raw_data %>%
-  # Clean column names
-  clean_names() %>%
-  
-  # Select the columns of interest for analysis
-  select(
-    address, 
-    application_for, 
-    ballots_cast, 
-    ballots_in_favour, 
-    ballots_opposed, 
-    close_date, 
-    poll_result, 
-    response_rate_met
-  ) %>%
-  
-  # Handle missing values by removing rows where address or ballots_cast are missing
-  filter(!is.na(address) & !is.na(ballots_cast)) %>%
-  
-  # Convert data types: dates and ballots
-  mutate(
-    close_date = as.Date(close_date, format = "%Y-%m-%d"),
-    ballots_cast = as.numeric(ballots_cast),
-    ballots_in_favour = as.numeric(ballots_in_favour),
-    ballots_opposed = as.numeric(ballots_opposed)
-  ) %>%
-  
-  # Add a new column for the percentage of ballots in favour
-  mutate(
-    percent_in_favour = (ballots_in_favour / ballots_cast) * 100
-  ) %>%
-  
[…]

### 2024-09-25 scripts/03-test_data.R

@@ -1,39 +1,46 @@
 #### Preamble ####
-# Purpose: Sanity check of the raw poll data
+# Purpose: Sanity check of the shelter occupancy data
 # Author: [Your Name]
 # Date: [Update to current date]
 # Contact: [Your Email]
 # License: MIT
-# Pre-requisites: Need to have raw poll data
+# Pre-requisites: Need to have raw shelter data
 # Any other information needed? None.
-
 
 #### Workspace setup ####
 library(tidyverse)
 
-
 #### Test data ####
 # Read in the raw data
-data <- read_csv("data/raw_data/raw_data.csv")
+data <- read_csv("data/analysis_data/analysis_data.csv")
 
 # Test for negative numbers in relevant numeric columns
-# Assuming ballots_cast, ballots_in_favour, and ballots_opposed should not have negative values
+# Checking service_user_count, occupancy_rate_beds, and capacity_actual_bed
 test_negative_values <- list(
-  ballots_cast = min(data$BALLOTS_CAST, na.rm = TRUE) >= 0,
-  ballots_in_favour = min(data$BALLOTS_IN_FAVOUR, na.rm = TRUE) >= 0,
-  ballots_opposed = min(data$BALLOTS_OPPOSED, na.rm = TRUE) >= 0
+  service_user_count = min(data$service_user_count, na.rm = TRUE) >= 0,
+  occupancy_rate_beds = min(data$occupancy_rate_beds, na.rm = TRUE) >= 0,
+  capacity_actual_bed = min(data$capacity_actual_bed, na.rm = TRUE) >= 0
 )
 
 # Display test results for negative values
 test_negative_values
 
 # Test for NAs in critical columns
-# Check if there are any NAs in columns like address, ballots_cast, and poll_result
+# Check if there are any NAs in columns like occupancy_date, service_user_count, and program_model
 test_na_values <- list(
-  address_na = all(!is.na(data$ADDRESS)),
-  ballots_cast_na = all(!is.na(data$BALLOTS_CAST)),
-  poll_result_na = all(!is.na(data$POLL_RESULT))
+  occupancy_date_na = all(!is.na(data$occupancy_date)),
+  service_user_count_na = all(!is.na(data$service_user_count)),
+  program_model_na = all(!is.na(data$program_model)),
+  sector_na = all(!is.na(data$sector)),
+  occupancy_rate_beds_na = all(!is.na(data$occupancy_rate_beds))
 )
 
 # Display test results for NAs
[…]

### 2024-09-25 paper/paper.qmd
Draft Complete, needs to do more editing

@@
 ---
-title: "Analysis of Public Opinion Polls Conducted by the City of Toronto"
-subtitle: "Insights on Front Yard Parking, Traffic Calming, and Boulevard Cafes"
+title: "Analyzing Homelessness and Shelter Use in Toronto: A Data-Driven Approach"
+subtitle: "Analyzing Homelessness and Shelter Use in Toronto: A Data-Driven Approach"
 author:
@@
 date-format: long
-abstract: "This paper analyzes data from the City of Toronto's public opinion polls on key municipal projects.
-We use a dataset covering several initiatives, including front yard parking, boulevard cafes, and traffic calming.
-Our analysis reveals significant variation in voter turnout and approval rates across these different projects.
-The findings provide insights into community engagement and policy preferences, with potential implications for future urban planning and public consultation processes"
+abstract: "This paper examines homelessness and shelter usage in Toronto by analyzing daily shelter occupancy data from the City of Toronto.
+Using data from 2021 to 2024, the analysis highlights key trends in shelter occupancy rates, bed capacity, and service user counts, broken down by program model and demographic sector.
+The results reveal a consistently high occupancy rate of beds, suggesting shelters are near capacity, with a growing demand for emergency services.
+Seasonal fluctuations in shelter usage, particularly during the winter months, point to the critical need for enhanced shelter services during extreme weather conditions.
+The analysis further uncovers disparities between emergency and transitional shelter capacities, emphasizing the necessity for targeted government policies to address the rising homelessness crisis.
+This paper aims to provide a deeper understanding of homelessness trends in Toronto and highlights the importance of data-driven solutions in addressing this ongoing social challenge."
 format: 
@@
 ---
-
-# Setup and Libraries
 
@@
 The state of being homeless is often understood as individuals' inability to stay in a permanent accommodation.
-As homelessness is a fundamental violation of several human rights, the United Nations points to state's failure to provide adequate and affordable housing as the leading cause of the staggering cases (<https://www.ohchr.org/en/special-procedures/sr-housing/homelessness-and-human-rights>).
-This phenomenon is also observed in Canada, where over 11% of Canadians have reported experiencing some form of homelessness (<https://www.statcan.gc.ca/o1/en/plus/5170-homelessness-how-does-it-happen>).
-Moreover, currently over 235,000 people experience homelessness everyday, according to statistics Canada (<https://www150.statcan.gc.ca/n1/pub/82-003-x/2021001/article/00002-eng.htm>).
+As homelessness is a fundamental violation of several human rights, the United Nations points to state's failure to provide adequate and affordable housing as the leading cause of the staggering cases (@ohchrhomelessness).
+This phenomenon is also observed in Canada, where over 11% of Canadians have reported experiencing some form of homelessness (@statcanhomelessness).
+Moreover, currently over 235,000 people experience homelessness everyday, according to statistics Canada (@statcanhealtharticle).
 While the Canadian state continues to brush over the severeness of the homeless crisis by restating their initiative and program, which frame the narrative in a extremely positive manner, we would like to investigate into the rapidly increasing homeless population to shed light on the need for a more proactive initiative from the state with housing supply and secure social safety net.
@@
 Additionally, since this date, the dataset includes occupancy data from hotel-based shelters operated by the Canadian Red Cross Society for refugee claimants.
-The raw dataset also includes fields such as the shelter group (e.g., adult, family, youth, refugee) and the organization operating the shelter.
+The raw dataset also includes fields such as the shelter group (e.g., adult, family, youth, refugee) and the organization operating the shelter(@torontodailyshelter2).
 
 ## Cleaned Data
+
+To prepare the dataset for analysis, several cleaning steps were undertaken.
+First, all missing values (NA) present in the columns occupancy_rate_beds and program_model were removed to ensure consistency across the key variables used in the analysis.
+This allowed for a clearer and more accurate understanding of trends related to bed occupancy rates and the types of shelter programs available.
+
+The dataset used spans from 2021 to the present, and no specific dates or years were filtered out during the cleaning process, as all data points within this range were deemed relevant.
+In terms of selecting the appropriate data for analysis, the focus was placed on variables that demonstrated the most significant trends over time, particularly in relation to recent news articles and reports concerning homelessness and shelter usage in Toronto.
+
+The variables kept in the cleaned data included those that track trends over time, such as occupancy_date, service_user_count, and program_model, as these were essential for understanding how shelter use and capacity evolved over the study period.
+
+Although no transformations or new variables were created, the data from each year was initially provided as separate files.
+These annual datasets were merged into a single file to streamline the analysis and make it easier to observe long-term trends.
+
+The cleaning process was conducted using tools and packages such as tidyverse, dplyr, janitor, and ggplot2, which facilitated efficient manipulation and visualization of the data.
+
+\newpage
 
@@
 #| label: tbl-observation-by-year
-#| fig-cap: Service User Count by year
 #| echo: false
@@
 # Table for observations by year
-kable(observations_table, caption = "Table: Number of observations by year")
+kable(observations_table, caption = "Number of observations by year")
 ```
@@
 #| label: tbl-summary-statistics
-#| fig-cap: Summary Statistics
 #| echo: false
@@
 
-kable(summary_stats, caption = "Table: Mean and standard deviation of key variables")
+kable(summary_stats, caption = "Mean and standard deviation of key variables")
 
@@
 # Results {#sec-results}
+
+## Tracking the Growth in Shelter Bed Occupancy Rates
 
@@
   geom_text(aes(label = round(Avg_Occupancy_Rate_Beds, 1)), vjust = -0.5, color = "black") +
-  labs(title = "Average Occupancy Rate of Beds by year",
-       x = "year", y = "Average Occupancy Rate (Beds)") +
+  labs(x = "year", y = "Average Occupancy Rate (Beds)") +
   theme_minimal() +
@@
 
-With the graph that detailed the average occupancy rate of beds from 2021 to 2024, it is evident that there is a increasing need for shelter–which most probably translates to an increasing homeless population in Toronto.
-Although the government initiative claimed that there are more resources allocated to assist......
+From 2021 to 2024, average shelter bed occupancy rates consistently exceeded 96%, with a notable increase to 98.2% by 2024 (Figure 2).
+This suggests that Toronto’s shelters are nearing full capacity year-round, reflecting both the rise in homelessness and insufficient shelter space.
+Despite claims of increased government funding for homelessness in 2023 (@thestarhomelessencampments), the data does not show a corresponding reduction in occupancy rates.
+Instead, this discrepancy highlights a lag between policy implementation and real-world outcomes.
 
 This result can also hint towards a the maximization of capacity that suggest the continuous need for resources, not a staggering increase, We will need to focus on other trends in data to get to the conclusion on whether there is a lack of resources in solving homelessness in Toronto.
+
+\newpage
+
+## Assessing Actual Bed Capacity by Program Type
 
@@
   geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "Actual Bed Capacity by Program Model and year",
-       x = "year", y = "Actual Bed Capacity") +
+  labs(x = "year", y = "Actual Bed Capacity") +
   theme_minimal()
@@
 The above graph shed light on the actual program model that exist in these shelters, giving us a deeper look into what resources the large homeless population in Toronto actually needs.
-The data is focused on two program models....
-each meaning....
+The data reveals a significant disparity between Emergency and Transitional shelter capacities.
+Emergency shelters, designed for short-term relief, saw consistent growth in capacity from 2021 to 2024, reflecting the city’s attempt to address immediate needs.
+Transitional shelters, which are critical for helping individuals move towards permanent housing, did not experience the same level of capacity expansion.
+This disparity suggests that while the city is focused on emergency crisis management, it is under-investing in long-term solutions to homelessness(@torontodailyshelter).
 
-with large disparity between the two types, we see a lack of desire in staying in shelter for an extended period of time, which may express the desire to get out of the homeless state or some undesirable nature of these shelters.
+With large disparity between the two types, we see a lack of desire in staying in shelter for an extended period of time, which may express the desire to get out of the homeless state or some undesirable nature of these shelters.
+Recent reports highlight that violence within Toronto’s shelter system has risen significantly, with incidents of interpersonal violence increasing by 283% between 2011 and 2021.
+This includes physical assaults, verbal abuse, and harassment, which create unsafe environments for both residents and staff.
+Overcrowding, lack of privacy, and restrictive policies further contribute to the discomfort many shelter users face.
+These conditions, coupled with the complex mental health and social challenges of residents, may explain why many prefer to leave the shelters as soon as possible.
+Improving safety and conditions within shelters is essential to making them more welcoming for longer-term stays and providing the necessary support for individuals to move out of homelessness permanently(@cbcsheltersafety).
 
-The gap between two capacities also can point to possible policy direction that the government and hone in on, as well as potential academic research on the underlying reason behind the great differences.
+The gap between the two capacities suggests potential policy directions for the government to focus on, such as increasing support for transitional shelters to provide more long-term solutions.
+This disparity could also prompt further research into the underlying causes, exploring whether funding limitations, policy focus, or other systemic factors are driving the imbalance.
 
@@
 With a growing occupancy rate adding on to the actual growth of capacity, there is a strong possibility that the homeless population is seeing a exponential growth that the state has to focus on solving–with a more assertive policy outlook.
+
+## Monthly Trends in Shelter User Counts
 
@@
 
-# Plot: Service User Count Over Months by year
 ggplot(monthly_data, aes(x = month, y = service_user_count, color = factor(year), group = year)) +
   geom_line(size = 1) +
-  labs(title = "Service User Count Over Months by year",
-       x = "month", y = "Service User Count", color = "Year") +
+  labs(x = "month", y = "Service User Count", color = "Year") +
   theme_minimal() +
@@
 
-After looking into capacities of homeless shelters in Toronto, we will now look into service user count from 2021 to 2024, with a monthly distribution that give us a clearer sense of how the capacities
+Having examined the capacities of homeless shelters in Toronto, we now turn to the service user count from 2021 to 2024.
+By analyzing the monthly distribution of service users, we gain a clearer understanding of the factors contributing to the gradual increase in emergency shelter capacity over time.
+
+Shelter use spiked dramatically during the winter months each year, reflecting the severe impact of extreme weather on Toronto's homeless population (Figure 4).
+Despite the city's efforts to expand shelter space by adding 400 new spaces and additional warming centers, the system continues to struggle with overwhelming demand.
+The seasonal fluctuation in service use points to the need for more robust winter preparedness, including additional warming centers, mobile shelters, or other emergency accommodations.
+By focusing on seasonal fluctuations, the city could better align its homelessness strategy with the predictable increase in demand during the colder months.
+(@cbctorontohomelesswinter).
+
+In response to the increasing demand for shelters, especially during the winter months, the data in Figure 4 clearly shows significant spikes in shelter usage from November to February each year.
+These peaks highlight the seasonal pressures on Toronto’s shelter system, with shelter capacity being consistently stretched during colder months.
+The introduction of additional winter-specific services, such as warming centers and new shelter spaces, may help alleviate some of the strain.
+However, the persistent upward trend in service user counts over the years underscores that even with these measures, the system continues to face overwhelming demand.
+The data indicates that without further expansion of both short-term and long-term housing solutions, the shelter system will remain under significant pressure.(@cbctorontowinterhomelessnessplan).
+
+## Demographic Breakdown of Shelter Users by Year
+
+Following the analysis of overall service user counts across the years, it is essential to examine the distribution of shelter usage by different demographic sectors, including men, women, mixed adults, and youth.
+The graph below presents the total service user count by sector for each year from 2021 to 2024, offering deeper insight into how shelter demand has varied among different population groups.
 
@@
   geom_bar(stat = "identity") +
-  labs(title = "Service User Count by sector for Each year", x = "sector", y = "Total Service User Count") +
+  labs(x = "sector", y = "Total Service User Count") +
   facet_wrap(~ year, ncol = 2) +
@@
 
-\newpage
+From first glance of the graph, we can see the large disparity between men and women organized in the few categories.
+According to research, by The National Women's Housing & Homelessness Network, based in Canada, average of 36% of homeless population are women.
+As we compare this data to the graph we see, service used by women do correlate proportionately with their share within the homeless population--while there might be bias within data collected.
+However, the same research found that 90% of the families using emergency shelters are headed by women--which can lead to a misscalculation in the actual number of women in need of a shelter.
+We could further theorize that if including women that use shelter within the family unit, the actual disparity between men and women might be smaller that the current number.
+National Women's Housing & Homelessness Network further suggested that in Canada, there are less shelter dedicated to women than men, men having more that double of dedicated shelters.
+This is a massive imporvenment that needs to be made in the gender disparity from the policy front, as well as a dataset better assist the imporving rights for homeless women.
+(@womenshomelessnesscanada)
 
-\appendix
+# Discussion
 
-# Appendix {.unnumbered}
+## Results and Implications
+
+The analysis of daily shelter occupancy data from 2021 to 2024 reveals several critical trends in Toronto’s shelter system.
+First, the bed occupancy rates consistently exceeded 96%, reaching a peak of 98.2% in 2024.
+This finding suggests that Toronto’s shelters are nearing full capacity year-round, reflecting both the rise in homelessness and the limited availability of shelter space.
+Despite efforts by the city, such as the addition of 400 new spaces, the shelter system remains strained, indicating an ongoing mismatch between demand and available resources.
+
+Seasonal variations in shelter use were also evident, with significant spikes during the colder months, particularly from November to February.
+This trend highlights the vulnerability of Toronto’s homeless population during extreme weather conditions, reinforcing the need for enhanced winter-specific services such as additional warming centers and emergency accommodations.
+
+The data also uncovers a disparity between emergency and transitional shelter capacities.
+While emergency shelters saw steady growth in capacity, transitional shelters did not experience the same level of investment.
+This gap reflects a policy focus on short-term crisis management at the expense of long-term solutions to homelessness.
+As a result, many individuals remain stuck in emergency shelters without adequate support to transition into permanent housing.
+Addressing this imbalance is critical for reducing the long-term impact of homelessness in Toronto.
+
+## Further Areas of Exploration
+
+Several areas warrant further exploration to fully understand the dynamics of homelessness and shelter use in Toronto.
+One key area is the demographic breakdown of shelter users, particularly the gender disparity in shelter access.
+While men dominate the use of emergency shelters, research suggests that women, especially those heading families, may be underrepresented in the data.
+Future studies should investigate the reasons behind this disparity, including whether women are more likely to seek shelter through family-focused programs, and the adequacy of shelter spaces dedicated to women.
+
+Additionally, the rising incidents of violence within shelters raise important questions about the safety and effectiveness of these facilities.
+With reports indicating a 283% increase in interpersonal violence in shelters over the past decade, further research is needed to explore the relationship between shelter conditions and users’ willingness to remain in these spaces(@cbcsheltersafety).
+Addressing safety concerns, overcrowding, and restrictive policies could make shelters more welcoming and supportive environments.
+
+Lastly, there is a need to explore the funding and policy limitations that contribute to the underinvestment in transitional shelters.
+By examining the systemic factors driving the imbalance between emergency and long-term shelter capacities, policymakers can develop more targeted solutions to address both immediate needs and long-term homelessness reduction.
 

### 2024-09-26 paper/paper.qmd
Update README and R scripts, remove unused files, rename project file
- Modified README.md and relevant R scripts
- Deleted unused model, datasheet, and replication files
- Renamed project file to toronto-shelter-data.Rproj
- Updated paper, references, and LLM usage files

@@
 title: "Analyzing Homelessness and Shelter Use in Toronto: A Data-Driven Approach"
-subtitle: "Analyzing Homelessness and Shelter Use in Toronto: A Data-Driven Approach"
+subtitle: "Understanding the Dynamics of Homelessness in Toronto through Data"
 author:
   - "Tim Chen"
-thanks: "Code and data are available at: https://github.com/timchen0326/sta304_paper_1."
+thanks: "Code and data are available at: https://github.com/timchen0326/toronto-shelter-data."
 date: today
@@
 
-# Load the processed data
 data <- read.csv(here("data/analysis_data/analysis_data.csv"))
@@
 
-The data used in this paper is derived from Open Data Toronto and is accessed through the opendatatoronto library (Gelfand 2022).
-The particular dataset analyzed here focuses on daily shelter and overnight services usage in Toronto, including information about shelter occupancy, capacity, and vacancy (Toronto 2024).
-All data analysis was conducted using R (R Core Team 2023) with the assistance of several packages: tidyverse (Wickham et al.
-2019), here (J.
-B.
-Kirill Müller 2020), dplyr (Hadley Wickham 2023), tibble (R.
-F.
-Kirill Müller, Hadley Wickham 2023), janitor (Sam Firke 2023), ggplot2 (Wickham 2016), and knitr (Xie 2023).
-
 The dataset is published by the City of Toronto and features daily shelter occupancy and capacity information across various shelter types, such as room-based and bed-based programs.
@@
 The raw dataset also includes fields such as the shelter group (e.g., adult, family, youth, refugee) and the organization operating the shelter(@torontodailyshelter2).
+
+The data for this paper was sourced from Open Data Toronto and accessed via the opendatatoronto library (@opendatatoronto).
+The dataset analyzed focuses on daily shelter and overnight service usage in Toronto, providing details on shelter occupancy, capacity, and vacancy (@torontodailyshelter2,).
+All data analysis was performed in R (@citeR) using a variety of packages, including tidyverse (@thereferencecanbewhatever), here (@here), dplyr (@dplyr), tibble (@tibble), janitor (Sam Firke 2023), ggplot2 (@ggplot2), and knitr (@knitr).
 
@@
 These annual datasets were merged into a single file to streamline the analysis and make it easier to observe long-term trends.
-
-The cleaning process was conducted using tools and packages such as tidyverse, dplyr, janitor, and ggplot2, which facilitated efficient manipulation and visualization of the data.
+@tbl-cleaned_data provides a sample of the cleaned data, while @fig-service-user-trend a scatter plot displaying all observations organized by the date of collection.
 
@@
 
-# Create the table
 kable(data_subset, col.names = c('Occupancy Date', 'Service User Count', 'Program Model'), 
@@
 
-# Create the scatter plot
 ggplot(data, aes(x = occupancy_date, y = service_user_count)) +
@@
 
-To gain a better understanding of the structure of the shelter occupancy data, we first examine the number of observations recorded in each year, as displayed in Table 1.
+To gain a better understanding of the structure of the shelter occupancy data, we first examine the number of observations recorded in each year, as displayed in @tbl-observation-by-year.
 Over the period from 2021 to 2024, the data collection appears relatively consistent, with the exception of 2024, which shows a noticeable decrease in the number of observations.
@@
 
-Moreover, Table 2 provides key summary statistics on critical variables such as occupancy rate (beds), service user count, and capacity (actual beds).
+Moreover, @tbl-summary-statistics provides key summary statistics on critical variables such as occupancy rate (beds), service user count, and capacity (actual beds).
 On average, the occupancy rate for beds is quite high at 96.18%, with a standard deviation of approximately 9.35%, indicating that most shelters are consistently near full capacity.
@@
 
-# Count the number of observations by year
 observations_by_year <- table(data$year)
-# Convert capacity_actual_bed to numeric (if it's not already)
 
-# Create a dataframe for the table
 observations_table <- data.frame(
@@
 
-# Calculate mean and standard deviation for relevant columns
 mean_occupancy_rate <- mean(data$occupancy_rate_beds, na.rm = TRUE)
@@
 
-# Table for observations by year
 kable(observations_table, caption = "Number of observations by year")
@@
 
-# Plot: Average Occupancy Rate of Beds by year
 ggplot(yearly_data, aes(x = factor(year), y = Avg_Occupancy_Rate_Beds)) +
@@
 
-From 2021 to 2024, average shelter bed occupancy rates consistently exceeded 96%, with a notable increase to 98.2% by 2024 (Figure 2).
+From 2021 to 2024, average shelter bed occupancy rates consistently exceeded 96%, with a notable increase to 98.2% by 2024 (@fig-occupancy-rate).
 This suggests that Toronto’s shelters are nearing full capacity year-round, reflecting both the rise in homelessness and insufficient shelter space.
@@
 This result can also hint towards a the maximization of capacity that suggest the continuous need for resources, not a staggering increase, We will need to focus on other trends in data to get to the conclusion on whether there is a lack of resources in solving homelessness in Toronto.
-
-\newpage
 
@@
 
-# Group by year, Program Model, and summarize actual bed capacity
 yearly_bed_capacity <- filtered_data %>%
@@
 
-# Plot: Actual Bed Capacity by Program Model and year
 ggplot(yearly_bed_capacity, aes(x = factor(year), y = Total_Bed_Capacity, fill = program_model)) +
@@
 
-The above graph shed light on the actual program model that exist in these shelters, giving us a deeper look into what resources the large homeless population in Toronto actually needs.
+@fig-bed-capacity shed light on the actual program model that exist in these shelters, giving us a deeper look into what resources the large homeless population in Toronto actually needs.
 The data reveals a significant disparity between Emergency and Transitional shelter capacities.
@@
 Overcrowding, lack of privacy, and restrictive policies further contribute to the discomfort many shelter users face.
-These conditions, coupled with the complex mental health and social challenges of residents, may explain why many prefer to leave the shelters as soon as possible.
+These circumstances, combined with the residents' complex mental health and social challenges, may explain why many prefer to leave the shelters promptly.
 Improving safety and conditions within shelters is essential to making them more welcoming for longer-term stays and providing the necessary support for individuals to move out of homelessness permanently(@cbcsheltersafety).
@@
 
-Shelter use spiked dramatically during the winter months each year, reflecting the severe impact of extreme weather on Toronto's homeless population (Figure 4).
+Shelter use spiked dramatically during the winter months each year, reflecting the severe impact of extreme weather on Toronto's homeless population (@fig-service-user-count).
 Despite the city's efforts to expand shelter space by adding 400 new spaces and additional warming centers, the system continues to struggle with overwhelming demand.
@@
 
-In response to the increasing demand for shelters, especially during the winter months, the data in Figure 4 clearly shows significant spikes in shelter usage from November to February each year.
+In response to the increasing demand for shelters, especially during the winter months, the data in @fig-service-user-count clearly shows significant spikes in shelter usage from November to February each year.
 These peaks highlight the seasonal pressures on Toronto’s shelter system, with shelter capacity being consistently stretched during colder months.
@@
 
-# Plot: Service User Count by sector for Each year
 ggplot(sector_data_filtered, aes(x = sector, y = Total_Service_User_Count, fill = sector)) +
@@
 
-From first glance of the graph, we can see the large disparity between men and women organized in the few categories.
+From first glance of @fig-service-sector, we can see the large disparity between men and women organized in the few categories.
 According to research, by The National Women's Housing & Homelessness Network, based in Canada, average of 36% of homeless population are women.
@@
 The analysis of daily shelter occupancy data from 2021 to 2024 reveals several critical trends in Toronto’s shelter system.
-First, the bed occupancy rates consistently exceeded 96%, reaching a peak of 98.2% in 2024.
+First, the bed occupancy rates consistently exceeded 96%, reaching a peak of 98.2% in 2024 (@fig-occupancy-rate).
 This finding suggests that Toronto’s shelters are nearing full capacity year-round, reflecting both the rise in homelessness and the limited availability of shelter space.
@@
 
-Seasonal variations in shelter use were also evident, with significant spikes during the colder months, particularly from November to February.
+Seasonal variations in shelter use were also evident, with significant spikes during the colder months, particularly from November to February (@fig-service-user-count).
 This trend highlights the vulnerability of Toronto’s homeless population during extreme weather conditions, reinforcing the need for enhanced winter-specific services such as additional warming centers and emergency accommodations.
 
-The data also uncovers a disparity between emergency and transitional shelter capacities.
+The data also uncovers a disparity between emergency and transitional shelter capacities (@fig-bed-capacity).
 While emergency shelters saw steady growth in capacity, transitional shelters did not experience the same level of investment.

### 2024-09-26 scripts/00-simulate_data.R

@@ -1,10 +1,10 @@
 #### Preamble ####
-# Purpose: Simulates data
+# Purpose: Simulates shelter occupancy data with random dates and various attributes
 # Author: Tim Chen
 # Date: 19 September 2024
 # Contact: timwt.chen@mail.utoronto.ca
-# Pre-requisites: None
-# Any other information needed? None
+# Pre-requisites: Install tidyverse and lubridate packages
+# Any other information needed? Data saved in CSV format for further use
 
 #### Workspace setup ####
 library(tidyverse)

### 2024-09-26 scripts/01-download_data.R

@@ -1,11 +1,11 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Downloads and saves the data from Toronto Open Data (Daily Shelter Overnight Service Occupancy and Capacity)
+# Author: Tim Chen
+# Date: 26 September 2024
+# Contact: timwt.chen@mail.utoronto.ca
+# Pre-requisites: Install opendatatoronto and tidyverse packages; access to the Toronto Open Data portal
+# Any other information needed? The data will be saved in CSV format for future use
+
 
 
 #### Workspace setup ####

### 2024-09-26 scripts/02-data_cleaning.R

@@ -1,12 +1,11 @@
 #### Preamble ####
-# Purpose: Cleans the raw poll data from Polls Conducted by the City dataset
-# Author: [Your Name]
-# Date: [Update to current date]
-#### Workspace setup ####
-# Contact: [Your Email]
-# License: MIT
-# Pre-requisites: Raw data should be available in 'data/raw_data/unedited_data.csv'
+# Purpose: Cleans the raw shelter occupancy data from Daily Shelter Overnight Service Occupancy and Capacity dataset
+# Author: Tim Chen
+# Date: 26 September 2024
+# Contact: timwt.chen@mail.utoronto.ca
+# Pre-requisites: Raw data should be available in 'data/raw_data/' folder with CSV files for each year from 2021 to 2024
 # Any other information needed? None
+
 
 #### Workspace setup ###
 # Load necessary libraries

### 2024-09-26 scripts/03-test_data.R

@@ -1,11 +1,11 @@
 #### Preamble ####
 # Purpose: Sanity check of the shelter occupancy data
-# Author: [Your Name]
-# Date: [Update to current date]
-# Contact: [Your Email]
-# License: MIT
-# Pre-requisites: Need to have raw shelter data
+# Author: Tim Chen
+# Date: 26 September 2024
+# Contact: timwt.chen@mail.utoronto.ca
+# Pre-requisites: Raw shelter data should be available in 'data/analysis_data/analysis_data.csv'
 # Any other information needed? None.
+
 
 #### Workspace setup ####
 library(tidyverse)

### 2024-09-26 paper/paper.qmd
Touch Ups and changing readme file

@@
 ggplot(data, aes(x = occupancy_date, y = service_user_count)) +
-  geom_jitter(color = "gray", alpha = 0.5, width = 0.2, height = 0.2) +
+  geom_jitter(color = "gray", alpha = 0.5, width = 0.2, height = 0.2, size = 0.1) + # Adjust size here
   theme_minimal() +
@@
 Similarly, the mean service user count stands at 38.41, with a high standard deviation of 35.41, suggesting that there is substantial variability in the number of individuals using the shelters across the dataset.
-Notably, the data for capacity (actual beds) is missing, which may indicate incomplete reporting or issues in data collection regarding shelter bed availability.
-Further analysis is required to understand the implications of this missing data.
 The 'Families' sector was excluded from the final visualization due to a significant portion of missing data in the occupancy rate, which resulted in unreliable service user counts.
@@
   geom_text(aes(label = round(Avg_Occupancy_Rate_Beds, 1)), vjust = -0.5, color = "black") +
-  labs(x = "year", y = "Average Occupancy Rate (Beds)") +
+  labs(x = "Year", y = "Average Occupancy Rate (Beds)") +
   theme_minimal() +
-  coord_cartesian(ylim = c(80, NA))
+  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
 
@@
   geom_bar(stat = "identity", position = "dodge") +
-  labs(x = "year", y = "Actual Bed Capacity") +
+  labs(x = "Year", y = "Actual Bed Capacity") +
   theme_minimal()
@@
   geom_line(size = 1) +
-  labs(x = "month", y = "Service User Count", color = "Year") +
+  labs(x = "Month", y = "Service User Count", color = "Year") +
   theme_minimal() +
@@
 #| echo: false
+library(scales)
+
 sector_data <- data %>%
@@
   geom_bar(stat = "identity") +
-  labs(x = "sector", y = "Total Service User Count") +
+  labs(x = "Sector", y = "Total Service User Count") +
   facet_wrap(~ year, ncol = 2) +
+  scale_y_continuous(labels = comma) +  # Format y-axis with whole numbers
+  scale_fill_discrete(name = "Sector") +  # Change legend title
   theme_minimal() +

### 2024-09-26 paper/paper.qmd
Fixes for citation

@@
 Additionally, since this date, the dataset includes occupancy data from hotel-based shelters operated by the Canadian Red Cross Society for refugee claimants.
-The raw dataset also includes fields such as the shelter group (e.g., adult, family, youth, refugee) and the organization operating the shelter(@torontodailyshelter2).
+The raw dataset also includes fields such as the shelter group (e.g., adult, family, youth, refugee) and the organization operating the shelter[@torontodailyshelter2].
 
-The data for this paper was sourced from Open Data Toronto and accessed via the opendatatoronto library (@opendatatoronto).
-The dataset analyzed focuses on daily shelter and overnight service usage in Toronto, providing details on shelter occupancy, capacity, and vacancy (@torontodailyshelter2,).
-All data analysis was performed in R (@citeR) using a variety of packages, including tidyverse (@thereferencecanbewhatever), here (@here), dplyr (@dplyr), tibble (@tibble), janitor (Sam Firke 2023), ggplot2 (@ggplot2), and knitr (@knitr).
+The data for this paper was sourced from Open Data Toronto and accessed via the opendatatoronto library [@opendatatoronto].
+The dataset analyzed focuses on daily shelter and overnight service usage in Toronto, providing details on shelter occupancy, capacity, and vacancy [@torontodailyshelter2].
+All data analysis was performed in R [@citeR] using a variety of packages, including tidyverse [@thereferencecanbewhatever], here [@here], dplyr [@dplyr], tibble [@tibble], janitor (Sam Firke 2023), ggplot2 [@ggplot2], and knitr [@knitr].
 
@@
 Transitional shelters, which are critical for helping individuals move towards permanent housing, did not experience the same level of capacity expansion.
-This disparity suggests that while the city is focused on emergency crisis management, it is under-investing in long-term solutions to homelessness(@torontodailyshelter).
+This disparity suggests that while the city is focused on emergency crisis management, it is under-investing in long-term solutions to homelessness[@torontodailyshelter].
 
@@
 These circumstances, combined with the residents' complex mental health and social challenges, may explain why many prefer to leave the shelters promptly.
-Improving safety and conditions within shelters is essential to making them more welcoming for longer-term stays and providing the necessary support for individuals to move out of homelessness permanently(@cbcsheltersafety).
-
+Improving safety and conditions within shelters is essential to making them more welcoming for longer-term stays and providing the necessary support for individuals to move out of homelessness permanently[@cbcsheltersafety].
 The gap between the two capacities suggests potential policy directions for the government to focus on, such as increasing support for transitional shelters to provide more long-term solutions.
@@
 By focusing on seasonal fluctuations, the city could better align its homelessness strategy with the predictable increase in demand during the colder months.
-(@cbctorontohomelesswinter).
+[@cbctorontohomelesswinter].
 
@@
 However, the persistent upward trend in service user counts over the years underscores that even with these measures, the system continues to face overwhelming demand.
-The data indicates that without further expansion of both short-term and long-term housing solutions, the shelter system will remain under significant pressure.(@cbctorontowinterhomelessnessplan).
+The data indicates that without further expansion of both short-term and long-term housing solutions, the shelter system will remain under significant pressure.[@cbctorontowinterhomelessnessplan].
 
@@
 This is a massive imporvenment that needs to be made in the gender disparity from the policy front, as well as a dataset better assist the imporving rights for homeless women.
-(@womenshomelessnesscanada)
+[@womenshomelessnesscanada]
 
@@
 Additionally, the rising incidents of violence within shelters raise important questions about the safety and effectiveness of these facilities.
-With reports indicating a 283% increase in interpersonal violence in shelters over the past decade, further research is needed to explore the relationship between shelter conditions and users’ willingness to remain in these spaces(@cbcsheltersafety).
+With reports indicating a 283% increase in interpersonal violence in shelters over the past decade, further research is needed to explore the relationship between shelter conditions and users’ willingness to remain in these spaces[@cbcsheltersafety].
 Addressing safety concerns, overcrowding, and restrictive policies could make shelters more welcoming and supportive environments.

### 2024-09-26 paper/paper.qmd
changing axis label

@@
   labs(x = "Year", y = "Actual Bed Capacity") +
+  scale_y_continuous(labels = scales::comma) +
   theme_minimal()
@@
   theme_minimal() +
-  scale_x_discrete(limits = month.abb)
+  scale_x_discrete(limits = month.abb) +
+  scale_y_continuous(labels = scales::comma)
+
 ```

### 2024-09-27 paper/paper.qmd
changing title

@@
 ---
-title: "Analyzing Homelessness and Shelter Use in Toronto: A Data-Driven Approach"
+title: "Toronto’s Shelter System Under Pressure: Investigating Homelessness and Bed Occupancy"
 subtitle: "Understanding the Dynamics of Homelessness in Toronto through Data"

### 2025-06-18 scripts/00-simulate_data.R
did download equally as badly

@@ -1,37 +1,0 @@
-#### Preamble ####
-# Purpose: Simulates shelter occupancy data with random dates and various attributes
-# Author: Tim Chen
-# Date: 19 September 2024
-# Contact: timwt.chen@mail.utoronto.ca
-# Pre-requisites: Install tidyverse and lubridate packages
-# Any other information needed? Data saved in CSV format for further use
-
-#### Workspace setup ####
-library(tidyverse)
-library(lubridate)
-
-#### Simulate data ####
-set.seed(304)
-
-# Define the start and end dates for random date generation
-start_date <- as.Date("2018-01-01")
-end_date <- as.Date("2023-12-31")
-
-# Number of random dates and observations
-n_dates <- 100
-
-# Create simulated data
-simulated_data <- tibble(
-  OCCUPANCY_DATE = as.Date(runif(n = n_dates, 
-                                 min = as.numeric(start_date), 
-                                 max = as.numeric(end_date)), 
-                           origin = "1970-01-01"),
-  SERVICE_USER_COUNT = rpois(n = n_dates, lambda = 150),
-  OCCUPANCY_RATE_BEDS = runif(n = n_dates, min = 0, max = 100),
-  CAPACITY_ACTUAL_BED = rpois(n = n_dates, lambda = 80),
-  PROGRAM_MODEL = sample(c("Emergency", "Transitional"), n_dates, replace = TRUE),
-  SECTOR = sample(c("Health", "Housing", "Education", "Community"), n_dates, replace = TRUE)
-)
-
-# Save the simulated data
-write_csv(simulated_data, "data/raw_data/simulated_data.csv")

### 2025-06-18 scripts/00-simulate_data.py

@@ -1,0 +1,39 @@
+#### Preamble ####
+# Purpose: Simulates shelter occupancy data with random dates and various attributes
+# Author: Tim Chen (translated to Python)
+# Date: 19 September 2024
+# Contact: timwt.chen@mail.utoronto.ca
+# Pre-requisites: pandas, numpy
+# Any other information needed? Data saved in CSV format for further use
+
+import pandas as pd
+import numpy as np
+
+# Set random seed for reproducibility
+np.random.seed(304)
+
+# Define date range
+start_date = pd.to_datetime("2018-01-01")
+end_date = pd.to_datetime("2023-12-31")
+
+# Number of observations
+n_dates = 100
+
+# Generate random dates
+random_dates = pd.to_datetime(
+    np.random.uniform(start_date.value, end_date.value, n_dates).astype('int64'),
+    origin='unix'
+)
+
+# Simulate data
+simulated_data = pd.DataFrame({
+    "OCCUPANCY_DATE": random_dates,
+    "SERVICE_USER_COUNT": np.random.poisson(lam=150, size=n_dates),
+    "OCCUPANCY_RATE_BEDS": np.random.uniform(0, 100, n_dates),
+    "CAPACITY_ACTUAL_BED": np.random.poisson(lam=80, size=n_dates),
+    "PROGRAM_MODEL": np.random.choice(["Emergency", "Transitional"], size=n_dates),
+    "SECTOR": np.random.choice(["Health", "Housing", "Education", "Community"], size=n_dates)
+})
+
+# Save to CSV
+simulated_data.to_csv("data/raw_data/simulated_data.csv", index=False)

### 2025-06-18 scripts/01-download_data.py
clean & test -- clean weird?

@@ -1,37 +1,29 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from Toronto Open Data (Daily Shelter Overnight Service Occupancy and Capacity)
-# Author: Tim Chen
-# Date: 26 September 2024
-# Contact: timwt.chen@mail.utoronto.ca
-# Pre-requisites: Install opendatatoronto and tidyverse packages; access to the Toronto Open Data portal
-# Any other information needed? The data will be saved in CSV format for future use
+# Purpose: Downloads and saves the data from the Open Data Toronto portal
+# Author: Amie Liu
+# Date: 16 January 2024
+# Contact: amie.liu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: None
 
+import pandas as pd
+from io import StringIO
+import requests
 
+base_url = "https://ckan0.cf.opendata.inter.prod-toronto.ca"
 
-#### Workspace setup ####
-library(opendatatoronto)
-library(tidyverse)
+url = base_url + "/api/3/action/package_show"
+params = { "id": "daily-shelter-overnight-service-occupancy-capacity"}
+package = requests.get(url, params = params).json()
 
-#### Download data ####
-# get package
-package <- show_package("21c83b32-d5a8-4106-a54f-010dbe49f6f2")
-package
- 
-# get all resources for this package
-resources <- list_package_resources("21c83b32-d5a8-4106-a54f-010dbe49f6f2")
- 
-# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
-datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
- 
-# load the first datastore resource as a sample
-data <- filter(datastore_resources, row_number()==1) %>% get_resource()
-data
+for idx, resource in enumerate(package["result"]["resources"]):
 
+       if resource["datastore_active"]:
+            url = base_url + "/datastore/dump/" + resource["id"]
+            resource_dump_data = requests.get(url)
 
+            # Writing raw data to CSV
+            raw_case_data = pd.read_csv(StringIO(resource_dump_data.text))
+            raw_case_data.to_csv("data/raw_data/daily-shelter-overnight-service-occupancy-capacity-2024.csvv")
+            
 
-#### Save data ####
-# [...UPDATE THIS...]
-# change the_raw_data to whatever name you assigned when you downloaded it.
[…]

### 2025-06-18 scripts/02-data_cleaning.R

@@ -1,67 +1,0 @@
-#### Preamble ####
-# Purpose: Cleans the raw shelter occupancy data from Daily Shelter Overnight Service Occupancy and Capacity dataset
-# Author: Tim Chen
-# Date: 26 September 2024
-# Contact: timwt.chen@mail.utoronto.ca
-# Pre-requisites: Raw data should be available in 'data/raw_data/' folder with CSV files for each year from 2021 to 2024
-# Any other information needed? None
-
-
-#### Workspace setup ###
-# Load necessary libraries
-library(readr)
-
-library(lubridate)
-library(tidyr)
-library(dplyr)
-library(janitor)
-
-# Read the CSV file
-df2021 <- read.csv("data/raw_data/daily-shelter-overnight-service-occupancy-capacity-2021.csv")
-df2022 <- read.csv("data/raw_data/daily-shelter-overnight-service-occupancy-capacity-2022.csv")
-df2023 <- read.csv("data/raw_data/daily-shelter-overnight-service-occupancy-capacity-2023.csv")
-df2024 <- read.csv("data/raw_data/daily-shelter-overnight-service-occupancy-capacity-2024.csv")
-
-# Convert the OCCUPANCY_DATE column to the desired date format
-df2021$OCCUPANCY_DATE <- as.Date(df2021$OCCUPANCY_DATE, format = "%y-%m-%d")
-df2022$OCCUPANCY_DATE <- as.Date(df2022$OCCUPANCY_DATE, format = "%y-%m-%d")
-df2023$OCCUPANCY_DATE <- as.Date(df2023$OCCUPANCY_DATE, format="%Y-%m-%dT%H:%M:%S")
-
-# Verify the changes
-head(df2021$OCCUPANCY_DATE)
-head(df2022$OCCUPANCY_DATE)
-
-# Save the updated dataframe back to CSV files
-write.csv(df2021, file = "data/raw_data/raw_data_2021.csv", row.names = FALSE)
-write.csv(df2022, file = "data/raw_data/raw_data_2022.csv", row.names = FALSE)
-write.csv(df2023, file = "data/raw_data/raw_data_2023.csv", row.names = FALSE)
-
-# Read the saved files
-file1 <- read.csv("data/raw_data/raw_data_2021.csv")
-file2 <- read.csv("data/raw_data/raw_data_2022.csv")
-file3 <- read.csv("data/raw_data/raw_data_2023.csv")
[…]

### 2025-06-18 scripts/02-data_cleaning.py

@@ -1,0 +1,64 @@
+#### Preamble ####
+# Purpose: Cleans the raw shelter occupancy data from Daily Shelter Overnight Service Occupancy and Capacity dataset
+# Author: Tim Chen (translated to Python)
+# Date: 26 September 2024
+# Contact: timwt.chen@mail.utoronto.ca
+# Pre-requisites: CSVs must exist in 'data/raw_data/' for 2021–2024
+
+import pandas as pd
+from pathlib import Path
+
+# Define input paths
+base_path = Path("data/raw_data")
+files = [
+    base_path / "daily-shelter-overnight-service-occupancy-capacity-2021.csv",
+    base_path / "daily-shelter-overnight-service-occupancy-capacity-2022.csv",
+    base_path / "daily-shelter-overnight-service-occupancy-capacity-2023.csv",
+    base_path / "daily-shelter-overnight-service-occupancy-capacity-2024.csv"
+]
+
+# Read and parse each file with appropriate date format
+df2021 = pd.read_csv(files[0], parse_dates=["OCCUPANCY_DATE"], date_format="%y-%m-%d", dayfirst=False)
+df2022 = pd.read_csv(files[1], parse_dates=["OCCUPANCY_DATE"], date_format="%y-%m-%d", dayfirst=False)
+df2023 = pd.read_csv(files[2], parse_dates=["OCCUPANCY_DATE"])
+df2023["OCCUPANCY_DATE"] = pd.to_datetime(df2023["OCCUPANCY_DATE"], format="%Y-%m-%dT%H:%M:%S")
+df2024 = pd.read_csv(files[3])
+
+# Save cleaned raw versions (optional but mirrors R script)
+df2021.to_csv(base_path / "raw_data_2021.csv", index=False)
+df2022.to_csv(base_path / "raw_data_2022.csv", index=False)
+df2023.to_csv(base_path / "raw_data_2023.csv", index=False)
+
+# Re-read the saved files
+file1 = pd.read_csv(base_path / "raw_data_2021.csv")
+file2 = pd.read_csv(base_path / "raw_data_2022.csv")
+file3 = pd.read_csv(base_path / "raw_data_2023.csv")
+file4 = pd.read_csv(base_path / "daily-shelter-overnight-service-occupancy-capacity-2024.csv")
+
+# Combine all data
+combined_data = pd.concat([file1, file2, file3, file4], ignore_index=True)
+
+# Save combined raw data
+combined_data.to_csv(base_path / "combined_data.csv", index=False)
+
[…]

### 2025-06-18 scripts/03-test_data.R

@@ -1,46 +1,0 @@
-#### Preamble ####
-# Purpose: Sanity check of the shelter occupancy data
-# Author: Tim Chen
-# Date: 26 September 2024
-# Contact: timwt.chen@mail.utoronto.ca
-# Pre-requisites: Raw shelter data should be available in 'data/analysis_data/analysis_data.csv'
-# Any other information needed? None.
-
-
-#### Workspace setup ####
-library(tidyverse)
-
-#### Test data ####
-# Read in the raw data
-data <- read_csv("data/analysis_data/analysis_data.csv")
-
-# Test for negative numbers in relevant numeric columns
-# Checking service_user_count, occupancy_rate_beds, and capacity_actual_bed
-test_negative_values <- list(
-  service_user_count = min(data$service_user_count, na.rm = TRUE) >= 0,
-  occupancy_rate_beds = min(data$occupancy_rate_beds, na.rm = TRUE) >= 0,
-  capacity_actual_bed = min(data$capacity_actual_bed, na.rm = TRUE) >= 0
-)
-
-# Display test results for negative values
-test_negative_values
-
-# Test for NAs in critical columns
-# Check if there are any NAs in columns like occupancy_date, service_user_count, and program_model
-test_na_values <- list(
-  occupancy_date_na = all(!is.na(data$occupancy_date)),
-  service_user_count_na = all(!is.na(data$service_user_count)),
-  program_model_na = all(!is.na(data$program_model)),
-  sector_na = all(!is.na(data$sector)),
-  occupancy_rate_beds_na = all(!is.na(data$occupancy_rate_beds))
-)
-
-# Display test results for NAs
-test_na_values
-
-# You can also add additional tests if needed, such as:
-# 1. Checking if occupancy rates are within a valid range (e.g., 0 to 100)
-test_valid_occupancy_rate_beds <- all(data$occupancy_rate_beds >= 0 & data$occupancy_rate_beds <= 100, na.rm = TRUE)
-
-# Display the result of the occupancy rate test
-test_valid_occupancy_rate_beds

### 2025-06-18 scripts/03-test_data.py

@@ -1,0 +1,46 @@
+#### Preamble ####
+# Purpose: Sanity check of the shelter occupancy data
+# Author: Tim Chen
+# Date: 26 September 2024
+# Contact: timwt.chen@mail.utoronto.ca
+# Pre-requisites: Raw shelter data should be available in 'data/analysis_data/analysis_data.csv'
+# Any other information needed? None.
+
+
+#### Workspace setup ####
+library(tidyverse)
+
+#### Test data ####
+# Read in the raw data
+data <- read_csv("data/analysis_data/analysis_data.csv")
+
+# Test for negative numbers in relevant numeric columns
+# Checking service_user_count, occupancy_rate_beds, and capacity_actual_bed
+test_negative_values <- list(
+  service_user_count = min(data$service_user_count, na.rm = TRUE) >= 0,
+  occupancy_rate_beds = min(data$occupancy_rate_beds, na.rm = TRUE) >= 0,
+  capacity_actual_bed = min(data$capacity_actual_bed, na.rm = TRUE) >= 0
+)
+
+# Display test results for negative values
+test_negative_values
+
+# Test for NAs in critical columns
+# Check if there are any NAs in columns like occupancy_date, service_user_count, and program_model
+test_na_values <- list(
+  occupancy_date_na = all(!is.na(data$occupancy_date)),
+  service_user_count_na = all(!is.na(data$service_user_count)),
+  program_model_na = all(!is.na(data$program_model)),
+  sector_na = all(!is.na(data$sector)),
+  occupancy_rate_beds_na = all(!is.na(data$occupancy_rate_beds))
+)
+
+# Display test results for NAs
+test_na_values
+
+# You can also add additional tests if needed, such as:
+# 1. Checking if occupancy rates are within a valid range (e.g., 0 to 100)
+test_valid_occupancy_rate_beds <- all(data$occupancy_rate_beds >= 0 & data$occupancy_rate_beds <= 100, na.rm = TRUE)
+
+# Display the result of the occupancy rate test
+test_valid_occupancy_rate_beds

### 2025-06-18 scripts/01-download_data.py
quick fixes

@@ -1,10 +1,9 @@
-#### Preamble ####
-# Purpose: Downloads and saves the data from the Open Data Toronto portal
-# Author: Amie Liu
-# Date: 16 January 2024
-# Contact: amie.liu@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: None
+# Purpose: Downloads and saves the data from Toronto Open Data (Daily Shelter Overnight Service Occupancy and Capacity)
+# Author: Tim Chen
+# Date: 26 September 2024
+# Contact: timwt.chen@mail.utoronto.ca
+# Pre-requisites: Install opendatatoronto and tidyverse packages; access to the Toronto Open Data portal
+# Any other information needed? The data will be saved in CSV format for future use
 
 import pandas as pd
 from io import StringIO

### 2025-06-18 scripts/03-test_data.py

@@ -1,46 +1,50 @@
 #### Preamble ####
 # Purpose: Sanity check of the shelter occupancy data
-# Author: Tim Chen
+# Author: Tim Chen (translated to Python)
 # Date: 26 September 2024
 # Contact: timwt.chen@mail.utoronto.ca
-# Pre-requisites: Raw shelter data should be available in 'data/analysis_data/analysis_data.csv'
-# Any other information needed? None.
+# Pre-requisites: pandas
+# Any other information needed? Expects data at 'data/analysis_data/analysis_data.csv'
 
+import unittest
+import pandas as pd
 
-#### Workspace setup ####
-library(tidyverse)
+class TestShelterOccupancyData(unittest.TestCase):
 
-#### Test data ####
-# Read in the raw data
-data <- read_csv("data/analysis_data/analysis_data.csv")
+    @classmethod
+    def setUpClass(cls):
+        cls.data = pd.read_csv("data/analysis_data/analysis_data.csv")
 
-# Test for negative numbers in relevant numeric columns
-# Checking service_user_count, occupancy_rate_beds, and capacity_actual_bed
-test_negative_values <- list(
-  service_user_count = min(data$service_user_count, na.rm = TRUE) >= 0,
-  occupancy_rate_beds = min(data$occupancy_rate_beds, na.rm = TRUE) >= 0,
-  capacity_actual_bed = min(data$capacity_actual_bed, na.rm = TRUE) >= 0
-)
+    # Test for non-negative values in numeric columns
+    def test_service_user_count_non_negative(self):
+        self.assertTrue((self.data["service_user_count"] >= 0).all())
 
-# Display test results for negative values
-test_negative_values
+    def test_occupancy_rate_beds_non_negative(self):
+        self.assertTrue((self.data["occupancy_rate_beds"] >= 0).all())
 
-# Test for NAs in critical columns
-# Check if there are any NAs in columns like occupancy_date, service_user_count, and program_model
-test_na_values <- list(
-  occupancy_date_na = all(!is.na(data$occupancy_date)),
-  service_user_count_na = all(!is.na(data$service_user_count)),
-  program_model_na = all(!is.na(data$program_model)),
-  sector_na = all(!is.na(data$sector)),
-  occupancy_rate_beds_na = all(!is.na(data$occupancy_rate_beds))
-)
+    def test_capacity_actual_bed_non_negative(self):
+        self.assertTrue((self.data["capacity_actual_bed"] >= 0).all())
 
[…]

### 2025-06-18 paper/paper.qmd
changed paper to python

@@
 
-```{r setup, include=FALSE}
+```{python setup, include=FALSE}
+#| include: false
 
-library(ggplot2)
-library(dplyr)
-library(lubridate)
-library(here)
-library(knitr)
+import numpy as np
+import pandas as pd
+import matplotlib.pyplot as plt
+from datetime import datetime
+from matplotlib.ticker import FuncFormatter
 
-data <- read.csv(here("data/analysis_data/analysis_data.csv"))
+data = pd.read_csv("../data/analysis_data/analysis_data.csv")
 
@@
 
-```{r, warning=FALSE}
+```{python, warning=FALSE}
 #| label: tbl-cleaned_data
+#| tbl-cap: Sample of Cleaned Shelter Occupancy Data
 #| echo: false
-data_head <- head(data)
 
-data_subset <- data_head[, c("occupancy_date", "service_user_count", "program_model")]
+# Subset and preview
+data_subset = data.loc[:, ["occupancy_date", "service_user_count", "program_model"]].head()
 
-kable(data_subset, col.names = c('Occupancy Date', 'Service User Count', 'Program Model'), 
-      align = "ccc", caption = "Sample of Cleaned Shelter Occupancy Data")
+# Rename columns for display
+data_subset.columns = ["Occupancy Date", "Service User Count", "Program Model"]
+
+data_subset
 
@@
 
-```{r, warning=FALSE}
+```{python, warning=FALSE}
 #| label: fig-service-user-trend
@@
 
-data$occupancy_date <- as.Date(data$occupancy_date, format="%Y-%m-%d")
+#prepare data
+data["occupancy_date"] = pd.to_datetime(data["occupancy_date"])
 
-ggplot(data, aes(x = occupancy_date, y = service_user_count)) +
-  geom_jitter(color = "gray", alpha = 0.5, width = 0.2, height = 0.2, size = 0.1) + # Adjust size here
-  theme_minimal() +
-  labs(x = "Date of Sample Collection", y = "Service User Count") +
-  scale_y_continuous(limits = c(0, 100)) +
-  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
+# Apply jitter manually
+x = data["occupancy_date"]
+y = data["service_user_count"]
+x_jittered = x + pd.to_timedelta(np.random.uniform(-0.2, 0.2, size=len(x)), unit="D")
+y_jittered = y + np.random.uniform(-0.2, 0.2, size=len(y))
+
+# Plot
+plt.figure(figsize=(10, 5))
+plt.scatter(x_jittered, y_jittered, color="gray", alpha=0.5, s=1)
+
+# Axis formatting
+plt.ylim(0, 100)
+plt.xlabel("Date of Sample Collection")
+plt.ylabel("Service User Count")
+plt.xticks(rotation=45)
+
+# Set yearly x-axis ticks
+years = pd.date_range(start=x.min(), end=x.max(), freq="YS")
+plt.xticks(years, [d.strftime("%Y") for d in years])
+
+plt.grid(False)
+plt.tight_layout()
+plt.show()
 
@@
 
-```{r, warning=FALSE}
+```{python, warning=FALSE}
 #| label: tbl-observation-by-year
+#| tbl-cap: Number of observations by year
+#| echo: false
+data["year"] = pd.to_datetime(data["occupancy_date"]).dt.year
+
+# Create summary table
+observations_table = (
+    data["year"]
+    .value_counts()
+    .sort_index()
+    .reset_index()
+    .rename(columns={"index": "Year", "year": "Number of Observations"})
+)
+
+# Display summary table
+observations_table
+```
+
+```{python, warning=FALSE}
+#| label: tbl-summary-statistics
+#| tbl-cap: Mean and standard deviation of key variables
 #| echo: false
 
-observations_by_year <- table(data$year)
+# Compute summary stats
+summary_stats = pd.DataFrame({
+    "Variable": [
+        "Occupancy Rate (Beds)",
+        "Service User Count",
+        "Capacity (Actual Beds)"
+    ],
+    "Mean": [
+        data["occupancy_rate_beds"].mean(),
+        data["service_user_count"].mean(),
+        data["capacity_actual_bed"].mean()
+    ],
+    "SD": [
+        data["occupancy_rate_beds"].std(),
+        data["service_user_count"].std(),
+        data["capacity_actual_bed"].std()
+    ]
+})
 
-observations_table <- data.frame(
-  Year = names(observations_by_year),
-  Number_of_Observations = as.vector(observations_by_year)
-)
+# Round for cleaner display (optional)
+summary_stats = summary_stats.round(5)
 
-mean_occupancy_rate <- mean(data$occupancy_rate_beds, na.rm = TRUE)
-sd_occupancy_rate <- sd(data$occupancy_rate_beds, na.rm = TRUE)
-
-mean_service_user_count <- mean(data$service_user_count, na.rm = TRUE)
-sd_service_user_count <- sd(data$service_user_count, na.rm = TRUE)
-
-mean_capacity_actual_bed <- mean(data$capacity_actual_bed, na.rm = TRUE)
-sd_capacity_actual_bed <- sd(data$capacity_actual_bed, na.rm = TRUE)
-
-kable(observations_table, caption = "Number of observations by year")
-```
-
-```{r, warning=FALSE}
-#| label: tbl-summary-statistics
-#| echo: false
-#| 
-summary_stats <- data.frame(
-  Variable = c("Occupancy Rate (Beds)", "Service User Count", "Capacity (Actual Beds)"),
-  Mean = c(mean_occupancy_rate, mean_service_user_count, mean_capacity_actual_bed),
-  SD = c(sd_occupancy_rate, sd_service_user_count, sd_capacity_actual_bed)
-)
-
-kable(summary_stats, caption = "Mean and standard deviation of key variables")
+# Display the table
+summary_stats
 
@@
 
-```{r, warning=FALSE}
+```{python, warning=FALSE}
 #| label: fig-occupancy-rate
@@
 
-yearly_data <- data %>%
-  group_by(year) %>%
-  summarise(Avg_Occupancy_Rate_Beds = mean(as.numeric(occupancy_rate_beds), na.rm = TRUE), 
-            .groups = "keep")
+# Group and compute average occupancy rate
+yearly_data = (
+    data.groupby("year")["occupancy_rate_beds"]
+    .mean()
+    .reset_index(name="Avg_Occupancy_Rate_Beds")
+)
 
-ggplot(yearly_data, aes(x = factor(year), y = Avg_Occupancy_Rate_Beds)) +
-  geom_bar(stat = "identity", fill = "skyblue") +
-  geom_text(aes(label = round(Avg_Occupancy_Rate_Beds, 1)), vjust = -0.5, color = "black") +
-  labs(x = "Year", y = "Average Occupancy Rate (Beds)") +
-  theme_minimal() +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
+# Plot bar chart
+plt.figure(figsize=(8, 5))
+bars = plt.bar(
+    yearly_data["year"].astype(str),
+    yearly_data["Avg_Occupancy_Rate_Beds"],
+    color="skyblue"
+)
 
+# Add labels above bars
+for bar in bars:
+    height = bar.get_height()
+    plt.text(bar.get_x() + bar.get_width() / 2, height + 1, f"{height:.1f}",
+             ha="center", va="bottom", color="black")
+
+# Axis labels and formatting
+plt.xlabel("Year")
+plt.ylabel("Average Occupancy Rate (Beds)")
+plt.ylim(0, yearly_data["Avg_Occupancy_Rate_Beds"].max() * 1.1)
+plt.tight_layout()
+plt.grid(False)
+plt.show()
 ```
@@
 
-```{r, warning=FALSE}
+```{python, warning=FALSE}
 #| label: fig-bed-capacity
@@
 #| echo: false
-filtered_data <- data %>%
-  filter(program_model %in% c("Emergency", "Transitional"))
 
-yearly_bed_capacity <- filtered_data %>%
-  group_by(year, program_model) %>%
-  summarise(Total_Bed_Capacity = sum(capacity_actual_bed, na.rm = TRUE), 
-            .groups = "keep")
+# Filter for Emergency and Transitional only
+filtered_data = data[data["program_model"].isin(["Emergency", "Transitional"])]
 
-ggplot(yearly_bed_capacity, aes(x = factor(year), y = Total_Bed_Capacity, fill = program_model)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  labs(x = "Year", y = "Actual Bed Capacity") +
-  scale_y_continuous(labels = scales::comma) +
-  theme_minimal()
+# Group and summarize bed capacity by year and program model
+yearly_bed_capacity = (
+    filtered_data
+    .groupby(["year", "program_model"])["capacity_actual_bed"]
+    .sum()
+    .reset_index(name="Total_Bed_Capacity")
+)
+
+# Prepare bar positions for grouped bars
+years = sorted(yearly_bed_capacity["year"].unique())
+models = ["Emergency", "Transitional"]
+bar_width = 0.35
+x = range(len(years))
+
+# Initialize plot
+fig, ax = plt.subplots(figsize=(8, 5))
+
+# Plot each program model as separate bars
+for i, model in enumerate(models):
+    subset = yearly_bed_capacity[yearly_bed_capacity["program_model"] == model]
+    offsets = [pos + (i - 0.5) * bar_width for pos in x]
+    ax.bar(offsets, subset["Total_Bed_Capacity"], width=bar_width, label=model)
+
+# Axes formatting
+ax.set_xticks(x)
+ax.set_xticklabels([str(y) for y in years])
+ax.set_xlabel("Year")
+ax.set_ylabel("Actual Bed Capacity")
+ax.legend(title="Program Model")
+ax.yaxis.set_major_formatter(FuncFormatter(lambda x, _: f"{int(x):,}"))  # Add commas
+plt.tight_layout()
+plt.grid(False)
+plt.show()
 
@@
 
-```{r , warning=FALSE}
+```{python, warning=FALSE}
 #| label: fig-service-user-count
@@
 #| echo: false
-monthly_data <- data %>%
-  group_by(year, month) %>%
-  summarise(service_user_count = sum(service_user_count, na.rm = TRUE), 
-            .groups = "keep")
 
-ggplot(monthly_data, aes(x = month, y = service_user_count, color = factor(year), group = year)) +
-  geom_line(size = 1) +
-  labs(x = "Month", y = "Service User Count", color = "Year") +
-  theme_minimal() +
-  scale_x_discrete(limits = month.abb) +
-  scale_y_continuous(labels = scales::comma)
+# Ensure occupancy_date is datetime and year/month columns exist
+data["occupancy_date"] = pd.to_datetime(data["occupancy_date"])
+data["year"] = data["occupancy_date"].dt.year
+data["month"] = data["occupancy_date"].dt.strftime("%b")  # e.g., Jan, Feb
+
+# Group by year and month, summing service_user_count
+monthly_data = (
+    data.groupby(["year", "month"])["service_user_count"]
+    .sum()
+    .reset_index()
+)
+
+# Set month order
+month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
+               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
+monthly_data["month"] = pd.Categorical(monthly_data["month"], categories=month_order, ordered=True)
+monthly_data = monthly_data.sort_values(["year", "month"])
+
+# Plot
+plt.figure(figsize=(10, 5))
+for year, group in monthly_data.groupby("year"):
+    plt.plot(group["month"], group["service_user_count"], label=year, linewidth=1)
+
+plt.xlabel("Month")
+plt.ylabel("Service User Count")
+plt.legend(title="Year")
+plt.grid(False)
+plt.tight_layout()
+plt.gca().yaxis.set_major_formatter(FuncFormatter(lambda x, _: f"{int(x):,}"))  # Comma formatting
+plt.show()
 
@@
 
-```{r, warning=FALSE}
+```{python, warning=FALSE}
 #| label: fig-service-sector
@@
 #| echo: false
-library(scales)
 
-sector_data <- data %>%
-  group_by(year, sector) %>%
-  summarise(Total_Service_User_Count = sum(service_user_count, na.rm = TRUE), .groups = "drop")
+import pandas as pd
+import matplotlib.pyplot as plt
+from matplotlib.ticker import FuncFormatter
 
-sector_data_filtered <- sector_data %>%
-  filter(sector != "Families")
+# Group data
+sector_data = (
+    data.groupby(["year", "sector"])["service_user_count"]
+    .sum()
+    .reset_index(name="Total_Service_User_Count")
+)
 
-ggplot(sector_data_filtered, aes(x = sector, y = Total_Service_User_Count, fill = sector)) +
-  geom_bar(stat = "identity") +
-  labs(x = "Sector", y = "Total Service User Count") +
-  facet_wrap(~ year, ncol = 2) +
-  scale_y_continuous(labels = comma) +  # Format y-axis with whole numbers
-  scale_fill_discrete(name = "Sector") +  # Change legend title
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  coord_cartesian(ylim = c(0, NA))
+# Filter out 'Families' if needed
+sector_data = sector_data[sector_data["sector"] != "Families"]
 
+# Consistent color map
+sector_colors = {
+    "Men": "#F8766D",
+    "Mixed Adult": "#7CAE00",
+    "Women": "#00BFC4",
+    "Youth": "#C77CFF"
+}
+
+# Define layout
+years = sorted(sector_data["year"].unique())
+ncols = 2
+nrows = -(-len(years) // ncols)
+
+fig, axes = plt.subplots(nrows=nrows, ncols=ncols, figsize=(12, 4 * nrows), constrained_layout=True)
+axes = axes.flatten()
+
+# Plot each year
+for i, year in enumerate(years):
+    ax = axes[i]
+    subset = sector_data[sector_data["year"] == year]
+    ax.bar(
+        subset["sector"],
+        subset["Total_Service_User_Count"],
+        color=[sector_colors.get(s, "gray") for s in subset["sector"]]
+    )
+    ax.set_title(str(year))
+    ax.set_xlabel("Sector")
+    ax.set_ylabel("Total Service User Count")
+    ax.set_xticks(range(len(subset)))
+    ax.set_xticklabels(subset["sector"], rotation=45, ha='right')
+    ax.yaxis.set_major_formatter(FuncFormatter(lambda x, _: f"{int(x):,}"))
+
+# Remove unused subplots
+for j in range(i + 1, len(axes)):
+    fig.delaxes(axes[j])
+
+# Add shared legend
+handles = [plt.Rectangle((0, 0), 1, 1, color=c) for c in sector_colors.values()]
+labels = list(sector_colors.keys())
+fig.legend(handles, labels, title="Sector", loc="center right")
+
+plt.show()
 ```
